Slovene NLTK Tagger

### 2012-01-06 paper/paper.tex
Paper, time spent for tagging, measurements graphs

\documentclass[11pt,a4paper,english,twocolumn]{article}
%\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{babel}
\usepackage{blindtext}
\usepackage{hyperref}

\title{\textbf{Slovene NLTK Tagger}}
\author{
Niko Colnerič \\
\footnotesize Faculty of Computer and \\
\footnotesize Information Science \\
\footnotesize University of Ljubljana \\
\and
Nejc Banič \\
\footnotesize Faculty of Computer and \\
\footnotesize Information Science \\
\footnotesize University of Ljubljana \\
}

\begin{document}
\maketitle
\thispagestyle{empty}
%\pagebreak

\section*{Abstract - NIKICC}
This paper describes the process of building Slovene tagger for use in NLTK.  
\section{Introduction - NIKICC}
Tagger processes a sequence of words and attaches a part of speech tag to each word. We implemented tagger for Slovene language in NLTK. NLTK stands for Natural Language Toolkit and is a library for use in Python (currently for version below 3, we used 2.7). NLTK already has some taggers for English and some other languages. Our goal was to implement NLTK tagger for Slovene language. The result of this project depends on two other larger projects. First is \textit{JOS}\cite{JOS}(in Slovene it stands for: "Jezikoslovno Označevanje Slovenskega jezika"), from which we used the corpuses (large and structured set of texts). Second, \textit{nltk-trainer}\cite{nltk-trainer}, is an open-source project hosting on Github, which was used for actual training of the tagger on Slovene sentences. Our work based mainly on putting all this pieces together into a working Slovene NLTK tagger.
\section{JOS corpora}
The basic component for building Slovene tagger is \textbf{JOS corpora}. It contains collections of various text in Slovenian language with annotation. For our project we used jos1M corpus that contain 1 million words with it's appropriate lemmas and morphosyntactic descriptions. The JOS is compatible with Slovene\textit{ MULTEXT-East morphosyntactic specifications Version 4}\cite{MULTEXT-East}. It basic purpose is to define word classes. For each class there are various attributes and their appropriate values, which they can be mapped into morphosyntactic descriptions (i.e. MSD). The structure of JOS corpora is in  \textbf{Extensible Markup Language (XML)}. We can easily manipulate XML files in various programming languages (e.g. Python with xml.dom.minidom). This is mandatory because \textit{nltk-trainer}\cite{nltk-trainer} expects special form of input file. For further information see ~\ref{Corpus transformation}.
\section{Nltk-trainer - NIKICC}
Nltk-trainer\cite{nltk-trainer} is project, whose author is Jacob Perkins, from which we took the code for building the tagger. We almost exclusively used the script \textit{train\_tagger.py}, which has the ability to generate NLTK taggers. Its mandatory argument is corpus in .pos format. Other optional arguments can choose wheather to generate a Sequential Tagger, a Brill Tagger or a Classifier Based Tagger. It also has the possibility to set default tag (the tag for words, for which the tagger doesn't know how to tag) and wheather to evaluate the tagger or not. For evaluation of the tagger (section \ref{Evaluation}) this argument was used. 
\section{Taggers description}
\subsection{Trigram tagger}
\textit{TODO - NIKICC}
\subsection{Brill tagger}
\textit{TODO - BANIČ}
\subsection{Naive Bayes tagger}
\textit{TODO - BANIČ}
\section{Implementation}
\subsection{Corpus transformation}
\textit{TODO - BANIČ}
\subsection{Training the tagger}
\textit{TODO - NIKICC}
\subsection{Usage with NLTK}
\textit{TODO - BANIČ}
\subsection{Encoding problems}
\textit{TODO - BANIČ}
\section{Evaluation}\label{Evaluation}
\textit{TODO - NIKICC}
\section{Conclusion}
\textit{TODO - BANIČ}

%\tableofcontents\newpage
%\blinddocument

%\thispagestyle{empty}
\begin{thebibliography}{99}
\bibitem{JOS} \url{http://nl.ijs.si/jos/}
\bibitem{nltk-trainer} \url{https://github.com/japerk/nltk-trainer}
\bibitem{MULTEXT-East} \url{http://nl.ijs.si/ME/V4/msd/html/msd-sl.html}
\end{thebibliography}
\end{document}

>>>>

\documentclass[10pt, conference, compsocconf]{IEEEtran}
%\hyphenation{op-tical net-works semi-conduc-tor}
\usepackage[utf8]{inputenc}
\usepackage[pdftex]{graphicx} % for images
\usepackage{hyperref}

\title{Slovene NLTK Tagger}
\author{
	Niko Colnerič \\
	\footnotesize Faculty of Computer and \\
	\footnotesize Information Science \\
	\footnotesize University of Ljubljana \\
	\footnotesize \texttt{niko.colneric@gmail.com} \\
	\and
	Nejc Banič \\
	\footnotesize Faculty of Computer and \\
	\footnotesize Information Science \\
	\footnotesize University of Ljubljana \\
	\footnotesize \texttt{MAIL@} \\
}

\begin{document}
\maketitle
\thispagestyle{empty}

\section*{Abstract} %NIKICC
This paper describes the process of building NLTK tagger for custom language. NLTK stands for Natural Language Toolkit, a library for use in Python (version 2.7).
Tagger is a object, which processes a sequence of words and attaches a part of speech tag to each word.
Our language of implemetation was Slovene.
We constructed sequential \textbf{Trigram tagger}, \textbf{Brill tagger} and classifier based \textbf{Naive Bayes tagger}, which differ in tagging accuracy and time consumed for tagging. All this differences are also detaily compared.
\\\\
\textit{Keywords:} Tagger, corpus, training

\section{Introduction} %NIKICC
Let us begin with the description of implemented taggers.

%NIKICC
\subsection[Trigram tagger]{Trigram tagger\footnote{Definition and images summerized from \cite{NLTKBOOK}}}
To understant how \textit{trigram} works, we should first understand \textit{unigram}.
Unigram taggers are based on a simple statistical algorithm: \textit{for each token, assign the tag that is most likely for that particular token}.
In the case of tagging with unigram tagger, we only consider the current token, in isolation from any larger context. The best we can do is tag each word with its \textit{a priori} most likely tag.
\par
An n-gram tagger is a generalization of a unigram\footnote{Unigram could also be reffered to as 1-gram} tagger whose context is the current word together with the part-of-speech tags of the $n-1$ preceding tokens, as shown in figure \ref{fig:trigram}.

\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.4\textwidth]{tag-context.png} 
\end{center}
\caption{Tagger Context}
\label{fig:trigram}
\end{figure}

The tag to be chosen, $t_{n}$, is circled, and the context is shaded in grey.
In the example of an n-gram tagger shown in figure \ref{fig:trigram}, we have $n=3$; that is, we consider the tags of the two preceding words in addition to the current word.
An n-gram tagger picks the tag that is most likely in the given context.
\par
Altrough we use the name trigram all throughtout this paper, we acctually refer to sequential tagger, which consists of
\textit{affix}\footnote{The Affix tagger learns prefix and suffix patterns to determine the part of speech tag for word.}, \textit{unigram}, \textit{bigram} and \textit{trigram} taggers in this order respectively. First we build affix, which is used as ad \textit{backoff-tagger} for unigram. Unigram is afterwards used as a backoff-tagger for bigram, which is finally used for trigram.

\subsection{Brill tagger} %BANIČ
\textit{TODO - BANIČ}
Banič glej ta link : nltk.googlecode.com/svn/trunk/doc/book/ch05.html
poglavje 5.6
\subsection{Naive Bayes tagger} %BANIČ
\textit{TODO - BANIČ}

\section{Related work}
\subsection{Nltk-trainer} %NIKICC
The code for building the tagger was taken form \textit{Nltk-trainer}\cite{nltk-trainer} project, whose author is Jacob Perkins.
We almost exclusively used the script \textit{train\_tagger.py}, which has the ability to generate NLTK taggers.
Its mandatory argument is a corpus in .pos format.
Other optional arguments can choose wheather to generate a Sequential Tagger, a Brill Tagger or a Classifier Based Tagger.
It also has the possibility to set default tag and wheather to evaluate the tagger or not.
This argument was used for evaluating constructed taggers (see section \ref{results}).

\subsection{JOS corpora} %BANIČ
The basic component for building Slovene tagger is \textbf{JOS corpora}.
It contains collections of various text in Slovenian language with annotation.
For our project we used jos1M corpus that contain 1 million words with it's appropriate lemmas and morphosyntactic descriptions.
The JOS is compatible with Slovene\textit{ MULTEXT-East morphosyntactic specifications Version 4}\cite{MULTEXT-East}.
It basic purpose is to define word classes.
For each class there are various attributes and their appropriate values, which they can be mapped into morphosyntactic descriptions (i.e. MSD).
The structure of JOS corpora is in  \textbf{Extensible Markup Language (XML)}.
We can easily manipulate XML files in various programming languages (e.g. Python with xml.dom.minidom).
This is mandatory because \textit{nltk-trainer}\cite{nltk-trainer} expects special form of input file.
For further information see \ref{Corpus transformation}.

\section{Implementation}
\subsection{Corpus transformation} %BANIČ
\label{Corpus transformation} 
\textit{TODO - BANIČ}
\subsection{Training the tagger} %NIKICC
\textit{TODO - NIKICC}
\subsection{Usage with NLTK} %BANIČ
\textit{TODO - BANIČ}
\subsection{Encoding problems} %BANIČ
\textit{TODO - BANIČ}

\section{Results} %NIKICC
\label{results}
For all constructed taggers, we here provide their accuracy. First we divided the learning set into two groups (\textit{fraction} is the procentage of sentances assigned to first group).
Now we construct the tagger on the first group and then evaluate its accuracy with the second group. The results, which were obtained by repeating this proces for various fractions, are shown in table \ref{tab:evaluation} and figure \ref{fig:evaluation}.

\begin{table}[htb]
\begin{center}
\begin{tabular}{c|c|c|c}
\textit{fraction} & \textit{Trigram} & \textit{NaiveBayes} & \textit{Brill} \\\hline\hline
0.75  &  0.8286  &  0.8451  &  0.8490\\
0.80  &  0.8304  &  0.8461  &  0.8514\\
0.82  &  0.8306  &  0.8465  &  0.8515\\
0.84  &  0.8299  &  0.8462  &  0.8508\\
0.86  &  0.8306  &  0.8472  &  0.8512\\
0.88  &  0.8299  &  0.8474  &  0.8517\\
0.90  &  0.8288  &  0.8467  &  0.8507\\
0.91  &  0.8278  &  0.8462  &  0.8489\\
0.92  &  0.8276  &  0.8464  &  0.8490\\
0.93  &  0.8281  &  0.8464  &  0.8491\\
0.94  &  0.8278  &  0.8461  &  0.8496\\
0.95  &  0.8289  &  0.8475  &  0.8510\\
0.96  &  0.8388  &  0.8538  &  0.8609\\
0.97  &  0.8401  &  0.8568  &  0.8612\\
0.98  &  0.8417  &  0.8588  &  0.8629\\
0.99  &  0.8379  &  0.8567  &  0.8578\\\hline\hline
\textit{average} & \textbf{0.8317} & \textbf{0.8490} & \textbf{0.8529}
\end{tabular}
\end{center}
\caption{Taggers accuracy for various fractions.}
\label{tab:evaluation}
\end{table}

\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.5\textwidth]{../evaluation/graph.pdf} 
\end{center}
\caption{Accuracy evaluation results from table \ref{tab:evaluation}.}
\label{fig:evaluation_speed}
\end{figure}

During this evaluation we discover an other paramater, which should be taken into account.
Under the assumption that most of the time was actually spent for tagging the sentances (besides just testing for equallity and percentage calculation are needed), we noticed that the time for evaluations differ drastically.
The results of resarching this observation are listed in table \ref{tab:evaluation_speed} and figure \ref{fig:evaluation_speed}. The difference is enormous. Brill and Trigram run at almost the same speed, while Naive Bayes runs much slower.

\begin{table}[htb]
\begin{center}
\begin{tabular}{c|c|c|c}
\textit{number of words} & \textit{Trigram} & \textit{NaiveBayes} & \textit{Brill} \\\hline\hline
   50  &  0.0003 &   4.1543  &  0.0011\\
   75  &  0.0004 &   6.2464  &  0.0014\\
  100  &  0.0006 &   8.5232  &  0.0018\\
  125  &  0.0008 &  10.6370  &  0.0020\\
  150  &  0.0009 &  12.4937  &  0.0024\\
  175  &  0.0011 &  14.5484  &  0.0027\\
  200  &  0.0011 &  16.3944  &  0.0029\\
  225  &  0.0013 &  18.6332  &  0.0032\\
  250  &  0.0015 &  20.7357  &  0.0035\\
  275  &  0.0017 &  22.8280  &  0.0041\\
  300  &  0.0018 &  24.8243  &  0.0042\\
  325  &  0.0018 &  27.4270  &  0.0043\\
  350  &  0.0022 &  29.2458  &  0.0051\\
  400  &  0.0023 &  32.9599  &  0.0052\\
  425  &  0.0024 &  36.0322  &  0.0056\\
  450  &  0.0027 &  38.1471  &  0.0062\\
  475  &  0.0027 &  39.2278  &  0.0064\\
  500  &  0.0029 &  41.2688  &  0.0063\\
\end{tabular}
\end{center}
\caption{Time spent in seconds for tagging different numbers of words.}
\label{tab:evaluation_speed}
\end{table}

\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.5\textwidth]{../evaluation/graph_speed.pdf} 
\end{center}
\caption{Time measurements from table \ref{tab:evaluation_speed}. NOTE: The green line is not visible because it is right under the red one.}
\label{fig:evaluation}
\end{figure}

\section{Conclusion} %BANIČ
\textit{TODO - BANIČ}

\section*{Acknowledgements}

\section{Currently not used}
The result of this project depends on two other larger projects. First is \textit{JOS}\cite{JOS}(in Slovene it stands for: "Jezikoslovno Označevanje Slovenskega jezika"), from which we used the corpuses (large and structured set of texts). Second, \textit{nltk-trainer}\cite{nltk-trainer}, is an open-source project hosting on Github, which was used for actual training of the tagger on Slovene sentences. Our work based mainly on putting all this pieces together into a working Slovene NLTK tagger.

\begin{thebibliography}{99}
\bibitem{JOS} \url{http://nl.ijs.si/jos/}
\bibitem{nltk-trainer} \url{https://github.com/japerk/nltk-trainer}
\bibitem{MULTEXT-East} \url{http://nl.ijs.si/ME/V4/msd/html/msd-sl.html}
\bibitem{NLTKBOOK} \url{http://www.nltk.org/book}
\end{thebibliography}
\end{document}
